### 2017-01-31 manuscript.rmd
incorporated multiple changes from co-authors

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew]. These challenges include non-independence of observations across species -- more closely related species will tend to be more similar to each other than distantly related species. Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG. Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on multiple pairwise comparisons (Fig. 1A) between contemporary traits that do not account for the fact that evolutionary relationships explain much of the structure of variation across species. This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

@@

The second CFG study we evaluate here is Levin *et al.* [@Levin:2016bp]. This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. They concluded that these animals exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Fig. 2). This contrasts with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw]. Levin *et al.* conclude that this provides biological justification for the concept of phyla and may provide a definition of phyla. We previously described concerns with their interpretations of these results [@Hejnol:2016hw]. Here we directly address the analyses themselves by examining the structure of the pairwise comparisons.

@@

> Figure 1 | Pairwise and phylogenetic comparative approaches. (a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs). This leads to many more comparisons than unique observations, making each comparison non-independent. (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches. Phylogenetic approaches are rarely used for functional genomics studies.

@@

Results in KMRR are based on pairwise comparisons (Figure 1A) between Tau within each gene family. They found the correlation of Tau for orthologs to be significantly greater than the correlation of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs. From this they concluded that their analyses support the ortholog conjecture. They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".

@@

When we reanalyze the data with phylogenetic independent contrasts (Fig. 1B), we did not find increased evolutionary change in expression following duplication events (Fig. XXKMRR). Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event. Of these, `r length(gene_trees_calibrated)` were successfully time calibrated. These calibrated pruned trees were used to calculate phylogenetic independent contrasts for Tau. They contain `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.

@@

> Figure XX_KMRR_Pairwise | This figures demonstrates that the pairwise analyses presented in Figure 2a of KMRR can be reproduced with the subset of data we consider here and our own independent code base. Each circle indicates the Tau Pearson correlation (rho) for a set of pairwise comparisons annotated with a specific node name of a given age and event type (*i.e.*, whether the divergence at that nodes was due to speciation or duplication). Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence.

@@

To test the hypothesis that distantly related paralogs arising from these older duplication events are the reason for the lower correlation of Tau between paralogs than between orthologs, we removed them. When we consider only the paralogs that arise from duplication events the same age or younger than the oldest speciation event, the observed pattern is very different. The correlation of Tau between these shallower paralog pairs, R=`r round( paralog_subset_r, 3)`, is much stronger than for the correlation between all paralog pairs, R=`r round( paralog_r, 3)`. This is much closer to the ortholog correlation, which has R=`r round( ortholog_r, 3) `.

@@

How changes in animal development relate to the evolution of animal diversity is a major question in evolutionary developmental biology (EvoDevo). To address this topic, Levin *et al.* [@Levin:2016bp] analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. As their title ("The mid-developmental transition and the evolution of animal body plans") indicates, they arrived at two major conclusions. First, animal development is characterized by a well-defined mid-developmental transition that marks the transition from an early phase of gene expression to a late stage of gene expression. Second, this transition helps explain the evolution of features observed among distantly related animals. Specifically, they concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases. Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development [@Kalinka:2010dw; @DomazetLoso:2010cn]. Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a definition of phyla.

@@

This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch. The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by canges along some of the same branches. This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree. The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other.

@@

Addressing the statistical non-independence of pairwise comparisons is not the only advantage of using phylogenetic comparative methods for functional genomics analyses. Another problem with the pairwise comparisons is that, except at the tips, they summarize changes along many branches in the phylogeny. Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pairwise comparison. Phylogenetic methods have the advantage of isolating the changes under consideration (Fig. 1b). The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches. There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss, but these omissions affect both methods.

>>>>

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew]. A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolve once in a common ancestor. This violates the fundamental assumption of independence of observations in standard statistical methods. Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG. Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on multiple pairwise comparisons (Fig. 1A) between contemporary traits that do not account for the fact that evolutionary relationships explain much of the structure of variation across species. This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

@@

The second CFG study we evaluate here is Levin *et al.* [@Levin:2016bp]. This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. Using pairwise comparisons (Figure 1A), they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases for these ten animals (Fig. XX). They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla and may provide an operational definition of phyla. We previously described concerns with these interpretations [@Hejnol:2016hw]. Here we directly address the analyses themselves by examining the structure of the pairwise comparisons.

@@

> Figure 1 | Pairwise and phylogenetic comparative approaches. (a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs). This leads to many more comparisons than unique observations, making each comparison dependent on others. (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches. Phylogenetic approaches are rarely used for functional genomics studies.

@@

Results in KMRR are based on pairwise comparisons (Figure 1A) between Tau within each gene family. They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs. From this they concluded that their analyses support the ortholog conjecture. They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".

@@

When we reanalyze the data with phylogenetic independent contrasts, we did not find increased evolutionary change in expression following duplication events (Fig. XXKMRR). Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event. Of these, `r length(gene_trees_calibrated)` were successfully time calibrated. These calibrated pruned trees were used to calculate phylogenetic independent contrasts for Tau. They contain `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.

@@

> Figure XX_KMRR_Pairwise | This figures demonstrates that the pairwise analyses presented in Figure 2a of KMRR can be reproduced with the subset of data we consider here and our own independent code base. Each circle indicates the Tau Pearson correlation (rho) for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs). Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence.

@@

To test the hypothesis that distantly related paralogs arising from these older duplication events (Supplementary Figure XXAscertainment a) are the reason for the lower correlation of Tau between paralogs than between orthologs, we removed them. When we consider only the paralogs that arise from duplication events the same age or younger than the oldest speciation event (Supplementary Figure XXAscertainment b), the observed pattern is very different. The correlation of Tau between these shallower paralog pairs, R=`r round( paralog_subset_r, 3)`, is much stronger than for the correlation between all paralog pairs, R=`r round( paralog_r, 3)`. This is much closer to the ortholog correlation, which has R=`r round( ortholog_r, 3) `.

@@

How changes in animal development relate to the evolution of animal diversity is a major question in evolutionary developmental biology (EvoDevo). To address this topic, Levin *et al.* [@Levin:2016bp] analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. They arrived at two major conclusions. First, animal development is characterized by a well-defined mid-developmental transition that marks the transition from an early phase of gene expression to a late stage of gene expression. Second, this transition helps explain the evolution of features observed among distantly related animals. Specifically, they concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases. Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development [@Kalinka:2010dw; @DomazetLoso:2010cn]. Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a definition of phyla.

@@

This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch. The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by canges along some of the same branches (Figure 1a). This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree. The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other.

@@

Addressing the statistical dependence of pairwise comparisons is not the only advantage of using phylogenetic comparative methods for functional genomics analyses. Another problem with the pairwise comparisons is that, except at the tips, they summarize changes along many branches in the phylogeny. Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pairwise comparison. Phylogenetic methods have the advantage of isolating the changes under consideration (Fig. 1b). The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches. There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss, but these omissions affect both methods.
